\protect A Web of Human Knowledge: Unravelling Wikipedia's First Link Network

### 2015-11-29 writeup/2015-11-wikipedia-network.tex
backup

We identified the set of first links directly referencing each article. 
Similar to the number of visits, we can examine the overall distribution and 
first links to understand the direct structure of taking one step through the 
FLN.

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/articles_ndegree.png}
  \caption{
    \textbf{Highest Ranking Articles by indegree}
  }
  We rank each article by the number of direct first links to the article (indegree). 
  \label{fig:ndegree list}

\end{figure*}
The highest ranking article by indegree is 
"United States" with 80,249 direct first links. Other high-ranking articles
include foundational abstract concepts such as "village," "species,"; 
sports associations such as "American Football," "Association Football"; 
and developed nations such as "France," "Japan," "Russia," "Australia," and 
the "Netherlands."

"Philosophy" and other philosophical concepts with many traversal visits
are not high-ranking by indegree.
"Philosophy" has an indegree of 581, with direct first links from articles about Philosophers and areas of Philosophy: "Existentialism and Humanism," "Predeterminism," "Synoptic Philosophy,"Qualia," "Dorothy Emmet," and "Christopher W. Morris."
While many articles accumulate at "Philosophy" (see traversal visits discussion), 
the accumulation is not the 
result of many articles directly referencing "Philosophy." 
Instead, the accumulation of first links, as we argue in our 
discussion of traversal visits, flows towards Philosophy as the 
ultimate anchor when genralizing from specific articles towards broader notions.

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/ndegree_loglog.png}
  \caption{
    \textbf{FLN Degree Distribution}
  }
  We construct a log-log plot fit our results to a linear model. The result is a 
  an excellent fit with r = -0.98, yielding a power law exponent of -0.79. 
  The distribution appears scale free with most articles having fewer than 9 
  direct first links, while few hold most.
  \label{fig:degree distribution}
\end{figure*}

As a distribution, we examine the degree of the articles in the FLN with at least
one direct first link. The average indegree 
is 3.6 direct links with a standard deviation of 89.5 links.
Only 4826 articles have more than 100 direct first links, $75\%$ have fewer than 
9. Overall the indegree appears scale-free with an r value of -0.98 as a linear log-log fit and a power law exponent of -0.79.

@@

We followed every possible path through the network, taking 232 million steps along the way to measure the accumulated number of traversal visits.

@@

In addition to identifying cycles and cycle lengths, we also measure the traversal path length, which includes articles outside
of cycles. 
Path length measures the number of links traversed until a repeated or invalid link as a means to guage the network's depth.
We discovered the longest path length is also 365, matching the longest cycle of Orthodox Liturgics. 
We also found similarly lengthy paths following the evolution of a place or topic through time: 
"1953 in Scotland" or "1560s Architecture", with articles sequentially proceeding by year, decade or era.

Of the 11 million articles, 5.5 million had an invalid link or linked back to the same article, yielding a path length of zero. 
The most common path length is 29, with an interquartile range (26, 30).
The median path length far exceeds the popular 6-degrees of separation (see for 
example the average number of academic publications separating scholars 
\cite{six_degrees}), indicating a greater network depth. The network depth suggests a large network of spanning a variety of ideas, which Wikipedia certainly is.
The distribution of path lengths is similarly scale-free with few articles at the extreme of 365 path lengths, while the majority 
is between 26-30.

\begin{figure}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/path_lengths_boxplot.png}
  \caption{
    \textbf{Path Length Distribution}
  }
  The median network depth measured by the number of first links in a path
  is 29. The 365 cycle for "Orthodox" liturgics is the outlier to the right, while
  other historical articles about "Scotland," "UK" and so on are slightly outside
  the third quartile. More than $75\%$ of articles have path lenghts between 
  $0$ and $50$ links.

@@

We can group articles lying on the same path to identify {\it basins}: 
a group of path-connected articles.
Since cycles identify only groups of articles with a closed set of links, 
we additionally measure and rank basins to capture groups of closely related
articles branching outside of a cycle into the rest of the FLN.
Akin to river networks, these basins are areas of FLN accumulation with a path 
flowing outwards to the rest of the FLN.
The highest ranking basins by the number of traversal visits are groups of articles
around "Philosophy" given the high number of traversal visits "Philosophy" has. 
However, restricting our attention to the 
 single articles around which basins formed,
 we find high ranking basins around similarly foundational ideas:
"Community", "Landmass", "Federal Government", "Presentation", and "Belief System". 
Around each article is a sets of related ideas. Around "community" are groups of articles expressing similar ideas: "Political Union," "State," and "Federation." 
These concepts naturally emerge from the First Link Network potentially indicating pillars, which 
anchor specific knowledge in a broader, simpler concept.

\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/basins.pdf}
  \caption{
    \textbf{basins around a single article}
  }
  \label{fig:basins}

@@

Measuring only traversal visits and indegree is limiting as it does not distinguish whether a particular article in a cycle 
is a funnel, directing many more paths inside the cycle than others. 
To distinguish among articles in a cycle, we also measured traversal funnels, or the number of 
paths an article directs towards a cycle. Here we count the number of traversal visits up to a cycle, 
so that accumulation does not flow to all articles in a cycle.
The importance here is to distinguish an article that happens to be connected to another article with many traversal visits 
from an article directly funneling many paths.

>>>>

We first identify the set of first links directly referencing a particular article 
by studying the indegree of the FLN. 
A higher indegree suggests more authors found the ideas in the article 
a pertinent direct reference. 
\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/articles_ndegree.png}
  \caption{
    \textbf{Highest Ranking Articles by indegree}
  }
  We rank each article by the number of direct first links to the article (indegree). The highest-ranking articles tend to represent abstraction for concepts
  comprised of various facets and parts.
  \label{fig:ndegree list}
\end{figure*}
We rank all 11 million articles by indegree to find 
the "United States" with 80,249 direct first links as the most referenced
Wikipedia article. 
Other high-ranking articles
include foundational abstract concepts such as "village," "species,"; 
sports associations such as "American Football," "Association Football"; 
and developed nations such as "France," "Japan," "Russia," "Australia," and 
the "Netherlands." These high ranking articles are useful abstractions: nations
describe a collection of individuals, a common culture, language, or 
geographical proximity; sports teams describe an ever changing collection of 
sports players often associated with a cultural identity and a geographical 
region. Since abstractions such as nations and teams are inherently comprised
of many parts, authors reference the larger abstraction when describing 
a part. For example, in describing the geographical region of north America, 
the United States, Canada, or Mexico are natural references anchoring the 
geographical location to a larger national identity.

"Philosophy" and other philosophical concepts with many traversal visits
are not among the highest-ranking articles by indegree.
"Philosophy" has an indegree of 581, with direct first links from articles about Philosophers and areas of Philosophy: "Existentialism and Humanism," "Predeterminism," "Synoptic Philosophy,"Qualia," "Dorothy Emmet," and "Christopher W. Morris."
While many articles accumulate at "Philosophy" (see traversal visits discussion), 
the accumulation is not the 
result of many articles directly referencing "Philosophy." 
Instead, the accumulation of first links, as we argue in our 
discussion of traversal visits, flows towards Philosophy as the 
ultimate anchor when genralizing from specific articles towards broader notions.
\begin{figure*}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/ndegree_loglog.png}
  \caption{
    \textbf{FLN Degree Distribution}
  }
  We construct a log-log plot fit our results to a linear model. The result is a 
  an excellent fit with r = -0.98, yielding a power law exponent of -0.79. 
  The distribution appears scale free with most articles having fewer than 9 
  direct first links, while few hold most.
  \label{fig:degree distribution}
\end{figure*}
The FLN's indegree exihibits a scale-free distribution where a few articles 
receive a most direct first references, while most articles receive few or none.
The average indegree for all 11 million articles is 3.6 direct first links with a standard deviation of 89.5 links.
Only 4826 articles have more than 100 direct first links and $75\%$ of articles
have fewer than 9. 
When fit to a linear model on a log-log scale (log(rank) versus log(indegree)), 
the model's r value is $-0.98$ suggesting a strong log-log linear fit 
with a power law exponent of $-0.79$.

\centerline{***}

Beyond indegree, our work expands the analysis of relations to understand 
how many more than two articles are organized and related. 
We develop the notion of a path to describe the structure of the FLN as a 
flow relating ideas. We traverse every possible path through the FLN, taking 
$232$ million steps along the way.

We first seek to describe the depth of the FLN: how many links does a 
connection of ideas span? We gauge the FLN's depth via path length or the number of links traversed until a repeated or invalid link (one outside the FLN). 
We discover the longest path length is 365
corresponding to the yearly calendar of Orthodox Liturgics.
Each day's Liturgics link to the next day's. Curiously, on the last calendar day, the last article simply links back to January 1, forming a 365-cycle 
(see discussion of cycles).
We also found similarly lengthy paths following the evolution of a place or topic through time: 
"1953 in Scotland" or "1560s Architecture", with articles sequentially proceeding by year, decade or era.
The longest paths connecting ideas are organized temporally.

\begin{figure}[tp!]
  \centering	
  \includegraphics[width=\columnwidth]{graphics/path_lengths_boxplot.png}
  \caption{
    \textbf{Path Length Distribution}
  }
  The median network depth measured by the number of first links in a path
  is 29. The 365 cycle for "Orthodox" liturgics is the outlier to the right, while
  other historical articles about "Scotland," "UK" and so on are slightly outside
  the third quartile. More than $75\%$ of articles have path lengths between 
  $0$ and $50$ links.

Of the 11 million articles, 5.5 million had an invalid link or linked back to the same article, yielding a path length of zero. 
This roughly corresponds to the official number of articles on Wikipedia 
$~4.7$ million as of November 2014---approximately half of the 11 million 
articles in the XML dump are redirects or disambiguations, not full articles.
The most common path length is 29, with an interquartile range (26, 30).
The median path length far exceeds the popular 6-degrees of separation (see for 
example the average number of academic publications separating scholars 
\cite{six_degrees}), indicating a greater network depth. The network depth suggests a large network of spanning a variety of ideas.
As a distribution, more than $75\%$ of articles have a path length below 
$50$ first links 
while a few temporally organized paths exceed 50 links 
(see figure~\ref{fig:Path Length Distribution}). 
How can we characterize the flow of ideas along a path? Next, we develop 
a metric to gauge the accumulation first link references
and characterize a global organization 
of ideas among articles.

@@

We followed every possible path through the network, incrementing
the number of traversal visits for every first link reference along a path 
to generate a total of 232 million traversal visits.

@@

@@

We can group articles lying on the same path to identify {\it basins}: 
a group of path-connected articles.
Since cycles identify only groups of articles with a closed set of links, 
we additionally measure and rank basins to capture groups of closely related
articles branching outside of a cycle into the rest of the FLN.
We rank basins by the total number of traversal visits in each of the articles
along the path forming the path.
Akin to river networks, these basins are areas of accumulation with a path 
flowing outwards to the rest of the FLN.

The highest ranking basins by the number of traversal visits are groups of articles
around "Philosophy." 
Since the accumulation of first links to "Philosophy" is so high, 
the paths leading to "Philosophy" carry many references.
The highest ranking paths include branches of philosophy flowing through 
"Awareness,"Existence," and "Consciousness" to "Philosophy." Other paths
include concepts around "Mathematics," scientific "Experiments," 
"Biology," and "Fact."
These paths link many specific articles to "Philosophy," each funneled through a particular domain.

Excluding basins around "Philosophy" we find other basins around 
foundational ideas such as "Community," "Landmass," "Federal Government," 
"Presentation," and "Belief System." 
The basins around each of these foundational notions are 
various paths containing related articles. For example around 
"Community" we find basins
flowing from "United States" to "Federal Republic" to "Political Union" to "State" culminating at "Community"; we also find basins flowing from 
"Public Policy" to "Executive (government)" to "Government" to State" and then 
to "Community"; we also find paths flowing through a similar chain beginning
with "Democracy," another beginning with "Constitution," another at 
"Dictatorship," and so on. The ideas build from specific means of organizing
a community (or society) then build up to "Community." 
Other basins around landmass for example begin at specific geographical regions
such as "Eastern Europe" building up to "Continent" and finally "Landmass."
Each basin accumulates first link references through various 
natural paths linking general themes to the foundational concept.

Which specific articles direct the path towards a particular concepts? 
Next we study the FLN using traversal funnels to understand the influence
a particular article exerts in directing the flow of first link references.

To distinguish articles exerting great influence 
To distinguish among articles in a cycle, we also measured traversal funnels, or the number of 
paths an article directs towards a cycle. Here we count the number of traversal visits up to a cycle, 
so that accumulation does not flow to all articles in a cycle.

@@

We also analyze article popularity for the highest ranking 1000 articles by 
traversal funnels. In October, the average page views per article is 
$22$ thousand with a standard deviation of $95$ thousand views. The distribution
is skewed with $75\%$ of articles reaching fewer than $20$ thousand page views. 
"Halloween" is the most popular article with $2.8$ million views in October,
likely a result of Halloween falling in the month of October. 
Other popular articles include "Scientology" ($463$ thousand views), "Clint Eastwood" 
($341$ thousand views),
and the "Cold War" ($298$ thousand views) although each has significantly fewer views compared to the views for "Halloween."
Other standout popular October articles include "24-hour" and "12-hour" clocks likely due
to Daylight savings in October and "Marriage" likely due to the drop in the number of
weddings in the months following October
\cite{weddings}.
"Philosophy" is ranks is the seventh most popular of the highest ranking articles by traversal funnels, but appears no where in the top 20 among the highest ranking
articles by traversal visits. "Philosophy" is a relatively popular article 
among articles influencing the shape of the network, but less popular among 
To distinguish among articles in a cycle, we also measured traversal funnels, or the number of 
paths an article directs towards a cycle. Here we count the number of traversal visits up to a cycle, 
so that accumulation does not flow to all articles in a cycle.
